Bayesian non-parametric inference for multivariate peaks-over-threshold models

### 2023-06-20 results.tex
blah

\begin{figure*}[htb]
    \centering
    % \caption{Energy score (top row) and Posterior Predictive Loss (bottom row) for various models on simulated data with ascending count of mixture components (indicated by plot heading) and number of dimensions (indicated by horizontal axis).  Gen indicates baseline.\label{fig:simpples}}
    \caption{Average energy score rise over baseline (on $\mathbb{S}_{\infty}^{d-1}$) for various models on simulated data, with ascending count of mixture components (indicated by plot heading) and number of dimensions (indicated by horizontal axis).  Note that pairwise betas is a moment-restricted model.\label{fig:simpples}}
    \includegraphics[height=3in, width = \textwidth]{./images/sim_es_rise}
\end{figure*}

\section{Data illustrations\label{sec:results}} 
We apply the aforementioned models to simulated angular data. We then consider the analysis
    of atmospheric data. To tackle the difficult problem of assessing the convergence 
    an MCMC chain for a large-dimensional model, we monitor the log-posterior 
    density. In all the examples considered MCMC samples produced stable tarces of the
    log-posterior in less than 40 thousand  iterations. Thus, so we use that as a burn-in, 
    and thereafter sample 10 thousand additional iterations. We then thin the chain by retaining one
    every ten samples, to obtain 1000 total samples. These are used to generate samples from the 
    posterior predictive densities. We used two different strategies implemment the MCMC samplers.
    For the models whose DP prior is centered around a log-normal distribution we used parallel 
    tempering. For the models gamma-centered models, we used a single chain. The temperature ladder was set as $t_s = 1.3^{s}$, 
    for $s \in \lbrace0,1,\ldots,5$. This was set empirically in order to produce
    acceptable swap probabilities both for the simulated data, and real data. Our attempts to use
    a single chain produced sample with very slow mixing. On the other hand, from our examples 
    we observe that posterior samples from the gamma-centered models, that do not use parallel 
    tempering, show slightly stronger autocorrelation than those of the log-normal centered 
    models. Parallel tempering produces chains with good mixing properties, but has a computational 
    cost that  grows linearly with the number of temperatures. In view of this, we use it for the

@@

\subsection{Simulation Study\label{subsec:simulated}}
To evaluate our proposed approach for angular measure estimation 
    we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$,
    for values of $d$ ranging between 2 and 32.  We generated each
    dataset as a mixture of multivariate log-normal distributions,
    projected onto $\mathbb{S}_{\infty}^{d-1}$. The generation 
    procedure is detailed in Algorithm~\ref{algo:simulated}.  For each
    model, the data is then projected onto the appropriate $\mathbb{S}_p^{d-1}$
    for model fitting--dirichlet and pairwise betas being fit on 
    $\mathbb{S}_1^{d-1}$, and the projected gamma models being fit on 
    $\mathbb{S}_{10}^{d-1}$.

\begin{algorithm}[h]  % algorithm using algorithm package
        \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}.
        $\mu_j$, $\Sigma_j$ are the parameters of the mixture component distribution; 
        $\pi$ is the probability vector assigning weight mixture components; $\delta_i$ 
        is the mixture component identifier for each simulated observation.}
        \begin{algorithmic}
        \For{$n_{\text{iter}}$ in $[1,\ldots,10]$}
        \For{$n_{\text{mix}}$ in $[1, 2, 4, 8]$}
            \For{$j$ in $1,\ldots,n_{\text{mix}}$}
                \State Generate $\mu_{j} \sim \mathcal{N}_{32}\left(\bm{0},\bm{I}\right)$
                \State Generate $\Sigma_{j}\sim\mathcal{IW}_{32}\left(70,70 \bm{I}\right)$
            \EndFor
            \State Generate $\pi\sim\text{Dirichlet}(\bm{10})$
            \For{$i$ in $1,\ldots,1000$}
                \State Generate $\delta_i \sim \text{Categorical}(\pi)$
                \State Generate $\bm{X}_i \sim \mathcal{LN}\left(\mu_{[\delta_i]},\Sigma_{[\delta_i]}\right)$
            \EndFor
            \For{$n_{\text{col}}$ in $[2,4,8,16,24,32]$}
                \State Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
            \EndFor
        \EndFor
        \EndFor
        \end{algorithmic}
    \end{algorithm}

Figure~\ref{fig:simpples} shows the average rise over baseline in energy score as
    calculated on $\mathbb{S}_{\infty}^{d-1}$ using the kernel metric described
    in Proposition~\ref{prop:rot}, for models trained on simulated data.  After training
    a model, a posterior predictive dataset is generated, and the energy score is
    calculated as a Monte Carlo approximation of Equation~\ref{eq:es}.
    The \emph{baseline} value is the energy score of a new dataset from the same
    generating distribution as the training dataset, evaluated against the training 
    dataset.  We considered a total of four variations of the Dirichlet-process mixture
    of projected gammas model: whether to restrict $\beta_{j\ell} := 1$ as the projected
    \emph{restricted} gamma (DP-PRG) akin to a projected Dirichlet, or allow them to be 
    learned with a gamma prior (DP-PG); and for the shape parameters whether to use a 
    gamma centering distribution (-G) or a log-normal centering distribution (-LN).
    For comparison we also included two other distributions on $\mathbb{S}_{1}^{d-1}$:
    The Dirichlet distribution (with gamma priors), and the pairwise betas distribution
    developed in~\cite{COOLEY2010}.  
    % Note that the pairwise betas model respects
    % the moment restriction generally assumed.\makenote{This phrasing needs to be changed.
    % The point I am trying to make is that because the moment-restriction assumption is 
    % discarded, a moment-restricted model is inadequate, as is borne out in the
    % simulation study.} \bruno{\bf I would avoid getting into the details of what the
    % pairwise beta model does. More important is to say why we chose it, and how available 
    % it is} 
    We observe that for simulated data using Algorithm~\ref{algo:simulated}, the 
    projected gamma models dominate, and within that class of model, the choice of 
    centering distribution is irrelevant.  For the projected \emph{restricted} 
    gamma models, the choice of a multivariate log-normal centering distribution 
    appears to negatively affect model fidelity.

@@

Our data corresponds to daily average values for the IVT magnitude
    along the coast of California.  The ERA-Interim data used covers the time period 
    1979 through 2014 (37 years) omitting leap days, and eight grid cells that 
    correspond to the coast of California.  The ERA5 data covers the time period 
    1979 through 2019 (42 years) with the same restriction, and  47 grid cells for 
    the coast of California.  This gives us the opportunity to illustrate the 
    performance of our method in multivariate settings of very different dimensions. 
    Figure~\ref{fig:gridlocs} provides a visual representation of the area these grid 
    cells represent.

@@

% Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy 
%     scores for the different models considered. We observe that, as in the case of the simulation 
%     study shown in Figures~\ref{fig:simpples}, the preferred model is the projected restricted 
%     gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
%     Interestingly, this preference for the restricted gamma form is the opposite of the result 
%     obtained in the simulation example.
Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy 
    scores for the different models considered. We observe that, contrary to the results in the
    simulation study in Figure~\ref{fig:simpples}, the preferred model is the projected restricted 
    gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
    Interestingly, this preference for the restricted gamma form is the opposite of the result 
    obtained in the simulation example.

>>>>

\begin{figure*}[htb]
    \centering
    \caption{Average energy score rise over baseline (on $\mathbb{S}_{\infty}^{d-1}$) for various 
    models fitted to simulated data, with ascending count of mixture components (indicated by plot
    heading) and number of dimensions (indicated by horizontal axis).  Note that pairwise betas is
    a moment-restricted model.\label{fig:simpples}}
    \includegraphics[height=3in, width = \textwidth]{./images/sim_es_rise}
\end{figure*}

\section{Data illustrations\label{sec:results}} 
We apply the aforementioned models to simulated angular data. We then consider the analysis
    of atmospheric data. To tackle the difficult problem of assessing the convergence 
    an MCMC chain for a large-dimensional model, we monitor the log-posterior 
    density. In all the examples considered, MCMC samples produced stable traces of the
    log-posterior in less than 40 thousand  iterations. We use that as a burn-in, 
    and thereafter sample 10 thousand additional iterations. We then thin the chain by retaining one
    every ten samples, to obtain 1000 total samples. These are used to generate samples from the 
    posterior predictive densities. We used two different strategies to implement the MCMC samplers.
    For the models whose DP prior is centered around a log-normal distribution we used parallel 
    tempering. This serves to overcome the very slow mixing that we observed for these cases.
    The temperature ladder was set as $t_s = 1.3^{s}$,  for $s \in \lbrace0,1,\ldots,5\rbrace$. This was 
    set empirically in order to produce acceptable swap probabilities both for the simulated data, 
    and real data. Parallel tempering produces chains with good mixing properties, but has a 
    computational cost that  grows linearly with the number of temperatures. Thus, for the 
    gamma-centered models, we used a single chain. We leverage the fast speed of each iteration, 
    to obtain a large number of samples, that are appropriately thinned to deal with a mild
    autocorrelation. In summary, the strategy for log-normal centered models is based on a costly
    sampler with good mixing properties. The strategy for the gamma-centered models is based on
    a cheap sampler that can be run for a large number of iterations.

@@

\subsection{Simulation Study\label{subsec:simulated}}
To evaluate our proposed approach for angular measure estimation 
    we consider simulated datasets on $\mathbb{S}_{\infty}^{d-1}$,
    for values of $d$ ranging between 2 and 32.  We generated each
    dataset as a mixture of multivariate log-normal distributions,
    projected onto $\mathbb{S}_{\infty}^{d-1}$. The generation 
    procedure is detailed in Algorithm~\ref{algo:simulated}.  We consider two gamma-centered
    and two log-normal centered DP mixture models, with and without restrictions in each
    case. We also consider a Dirichlet distribution and 
    a pairwise betas model \citep{COOLEY2010}, fitted using the \verb|BMAmevt| 
    package in R~\citep{BMAmevt}. For the DP mixture
    models, the data are projected onto $\mathbb{S}_{10}^{d-1}$. For the other
    two models they are projected on $\mathbb{S}_1^{d-1}$.

\begin{algorithm}[ht]  % algorithm using algorithm package
        \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}.
        $\mu_j$, $\Sigma_j$ are the parameters of the mixture component distribution; 
        $\pi$ is the probability vector assigning weight mixture components; $\delta_i$ 
        is the mixture component identifier for each simulated observation.}
        \begin{algorithmic}
        \For{$n_{\text{iter}}$ in $[1,\ldots,10]$}
        \For{$n_{\text{mix}}$ in $[1, 2, 4, 8]$}
            \For{$j$ in $1,\ldots,n_{\text{mix}}$}
                \State Generate $\mu_{j} \sim \mathcal{N}_{32}\left(\bm{0},\bm{I}\right)$
                \State Generate $\Sigma_{j}\sim\mathcal{IW}_{32}\left(70,70 \bm{I}\right)$
            \EndFor
            \State Generate $\pi\sim\text{Dirichlet}(\bm{10}_{n_{\text{mix}}})$
            \For{$i$ in $1,\ldots,1000$}
                \State Generate $\delta_i \sim \text{Categorical}(\pi)$
                \State Generate $\bm{X}_i \sim \mathcal{LN}\left(\mu_{[\delta_i]},\Sigma_{[\delta_i]}\right)$
            \EndFor
            \For{$n_{\text{col}}$ in $[2,4,8,16,24,32]$}
                \State Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
            \EndFor
        \EndFor
        \EndFor
        \end{algorithmic}
    \end{algorithm}

Figure~\ref{fig:simpples} shows the average rise over baseline in energy score as
    calculated on $\mathbb{S}_{\infty}^{d-1}$ using the kernel metric described
    in Proposition~\ref{prop:rot}, for models trained on simulated data.  After training
    a model, a posterior predictive dataset is generated, and the energy score is
    calculated as a Monte Carlo approximation of Equation~\ref{eq:es}.
    The \emph{baseline} value is the energy score of a new dataset from the same
    generating distribution as the training dataset, evaluated against the training 
    dataset.  
%    We considered a total of four variations of the Dirichlet-process mixture
%    of projected gammas model: whether to restrict $\beta_{j\ell} := 1$ as the projected
%    \emph{restricted} gamma (DP-PRG) akin to a projected Dirichlet, or allow them to be 
%    learned with a gamma prior (DP-PG); and for the shape parameters whether to use a 
%    gamma centering distribution (-G) or a log-normal centering distribution (-LN).
%    For comparison we also included two other distributions on $\mathbb{S}_{1}^{d-1}$:
%    The Dirichlet distribution (with gamma priors), and the pairwise betas distribution
%    developed in~\cite{COOLEY2010}.  
    % Note that the pairwise betas model respects
    % the moment restriction generally assumed.\makenote{This phrasing needs to be changed.
    % The point I am trying to make is that because the moment-restriction assumption is 
    % discarded, a moment-restricted model is inadequate, as is borne out in the
    % simulation study.} \bruno{\bf I would avoid getting into the details of what the
    % pairwise beta model does. More important is to say why we chose it, and how available 
    % it is}
    \makenote{It was anticipated that the log-normal centering distribution would place
    more mass of the  DP into regions of the support with active clusters, which would 
    generate new clusters more readily.  For the simulated data,} 
    we observe that the projected gamma models dominate, and within that class
    of model, the choice of centering distribution is immaterial.  For the 
    projected \emph{restricted} gamma models, the choice of a multivariate log-normal 
    centering distribution appears to \makenote{somewhat} negatively affect model 
    fidelity.  \makenote{Additionally, the log-normal models with the additional computational cost
    of multivariate normal evaluation and parallel tempering, take approximately six times
    longer to sample relative to the gamma models.}
    % \bruno{The comparison needs some information about the computing times.}

@@

Our data correspond to daily average values for the IVT magnitude
    along the coast of California.  The ERA-Interim data used covers the time period 
    1979 through 2014 (37 years) omitting leap days, and eight grid cells that 
    correspond to the coast of California.  The ERA5 data cover the time period 
    1979 through 2019 (42 years) with the same restriction, and  47 grid cells for 
    the coast of California.  This gives us the opportunity to illustrate the 
    performance of our method in multivariate settings of very different dimensions. 
    Figure~\ref{fig:gridlocs} provides a visual representation of the area these grid 
    cells cover.

@@

% Table~\ref{tab:dev} shows the values of the estimated posterior predictive loss and energy 
%     scores for the different models considered. We observe that, as in the case of the simulation 
%     study shown in Figures~\ref{fig:simpples}, the preferred model is the projected restricted 
%     gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
%     Interestingly, this preference for the restricted gamma form is the opposite of the result 
%     obtained in the simulation example.
Table~\ref{tab:dev} shows the values of the estimated energy 
    scores for the different models considered. We observe that, contrary to the results in the
    simulation study in Figure~\ref{fig:simpples}, the preferred model is the projected restricted 
    gamma models, though for the lower-dimensional ERA-Interim data, all models perform comparably.
